---
title: "Assignment 2"
author: "Group A -"
date: ""
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1 Findings

+ What are the major determinants of high crime rates?
+ Are there areas of the US with unusually low or high crime rates that do not conform to the general pattern?
+ Are the causes of violent and non-violent crime similar, or are there important differences?

\newpage

## 2 Statistical methodology

### Cleaning Data

After the exploratory data analysis completed assignment 1 we decided to do the following to clean up the data:  

+ Remove the NAs in medIncome and pctEmploy (Figure 1)
+ Remove unexpected values in ownHousMed, ownHousQrange, rentMed and rentQrange (Figure 1)
+ Change the Pacific region to West (Figure 2)
+ Use pctUrban values to create a rural/urban variable and make it a factor (Figure 3)

\newpage


## 3 Appendix  

Packages
```{r, results='hide', message=FALSE}
library(dplyr)
load("~/Documents/University/Year3/ST404/Assignment2/USACrime.Rda")
```

Figure 1 - Removing unexpected values and NA values
```{r}
USACrime$ownHousMed[USACrime$ownHousMed==500001] <- NA
USACrime$ownHousQrange[USACrime$ownHousQrange==0] <- NA
USACrime$rentMed[USACrime$rentMed==1001] <- NA
USACrime$rentQrange[USACrime$rentQrange==0] <- NA
USACrime.2 <- na.omit(USACrime)
```

Figure 2 - Changing Pacific region to West
```{r}
USACrime.2$region[USACrime$region.2 == "Pacific"] <- "West"
```

Figure 3 - Changing the pctUrban variables to factor
```{r}
USACrime.2 <- mutate(USACrime.2, isUrban = if_else(pctUrban < 85, "Not Urban", "Urban")) 
USACrime.2$isUrban <- as.factor(USACrime.2$isUrban)
```

